Mental Health Survey Data Cleaning and Analysis 


---
title: "Mental Health Survey Data Cleaning and Analysis"
author: "QUAZI NASHRA AHMAD"
date: "06-22-2024"
output: html_document
---

```{r setup8, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup, include=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyverse)

```

#Justification
Missing Values
Empty Strings to NA: Empty strings do not provide any information and should be treated as missing values.
Specific "NA" Values: Sometimes, "NA" might be inputted as a string instead of being left blank. These need to be converted to proper NA values for consistency.
Dropping Rows with Missing Age: Age is crucial for this analysis. Without it, the row cannot contribute meaningfully to our study of age distribution.
Outliers and Errors
Age Range (18-100): This range is chosen based on typical age expectations for survey respondents. Values outside this range are likely errors or outliers that can distort the analysis.
Data Type Conversion: Ensuring correct data types allows for accurate computations and visualizations. For example, treating Age as numeric enables arithmetic operations and histogram creation.
Visualization
Histogram with Bin Width 5: This bin width provides a balance between granularity and readability, showing detailed distribution without being overly fragmented.
Bar Graph for Employment Status and Family History: This bar graph allows us to compare the counts of respondents based on their employment status and whether they have a family history of mental illness. By using different colors for family history, we can easily see the distribution and relationship between these two factors.


```{r setup1, include=FALSE}
# Load the data
getwd()  # Check current working directory
setwd("C:/Users/Naser/Desktop/QNA")  # Set new working directory
data <- read.csv("survey.csv")
```

#We select only the first 5 columns that are most relevant for this analysis: Age, Gender, Country, self_employed, and family_history.

```{r setup2,include=FALSE}
# Display the structure of the data
str(data[, c("Age", "Gender", "Country", "self_employed", "family_history")])

# Display the first few rows of the selected columns
head(data[, c("Age", "Gender", "Country", "self_employed", "family_history")])
```

#We convert empty strings to NA for easier handling of missing data. We specifically handle columns where "NA" might appear as a string value, converting them to proper NA values.



```{r setup3, include=FALSE}
# Select relevant columns
data <- data %>% select(Age, Gender, Country, self_employed, family_history)

# Convert empty strings to NA
data[data == ""] <- NA

# Handle specific missing values
cols_with_na <- c("self_employed", "family_history")
data[cols_with_na] <- lapply(data[cols_with_na], function(x) ifelse(x == "NA", NA, x))

#We ensure that each column is of the correct data type: Age as numeric, and others as factors. This is crucial for proper analysis and visualization.

# Convert relevant columns to appropriate data types
data$Age <- as.numeric(data$Age)
data$Gender <- as.factor(data$Gender)
data$Country <- as.factor(data$Country)
data$self_employed <- as.factor(data$self_employed)
data$family_history <- as.factor(data$family_history)

```

```{r setup4, include=FALSE}
# Summary of cleaned data
summary(data)
```

#We create a histogram to visualize the distribution of Age among survey respondents. The bin width is set to 5 for a clear and detailed distribution.


```{r echo=FALSE}
ggplot(data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Age Distribution of Survey Respondents",
       x = "Age",
       y = "Count")

```

#We create a bar graph to visualize the distribution of gender among survey respondents. The bin width is set to 0.7 for a clear and detailed distribution.

```{r echo=FALSE}
ggplot(data, aes(x = Gender)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Gender Distribution of Survey Respondents",
       x = "Gender",
       y = "Count")
```

##We create a bar graph to visualize the distribution of country among survey respondents. The bin width is set to 0.7 for a clear and detailed distribution.

```{r echo=FALSE}

ggplot(data, aes(x = Country)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.7) +
  theme_minimal() +
  coord_flip() +
  labs(title = "Country Distribution of Survey Respondents",
       x = "Country",
       y = "Count")
```

##We create a bar graph to visualize the distribution of self employed among survey respondents. The bin width is set to 0.7 for a clear and detailed distribution.

```{r echo=FALSE}
ggplot(data, aes(x = self_employed)) +
  geom_bar(fill = "blue", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Self-Employed Distribution of Survey Respondents",
       x = "Self-Employed",
       y = "Count")
```
